---
title: "Calculating Zooplankton Metrics & MMI"
author: "Karen BLocksom"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Zooplankton Metrics & MMI}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Preliminaries

  One of the primary goals of this package is to calculate zooplankton metrics and the zooplankton MMI as used in the National Lakes Assessment (NLA) conducted in 2012 and 2017 by EPA's Office of Water. Two zooplankton samples were collected at sites, one with a coarse mesh net (ZOCN sample type) and one with a fine mesh net (ZOFN). Samples were processed in the laboratory to obtain the taxon name, number of individuals, and biomass information. Using the tow volume for each sample, densities, counts, and biomass were calculated for each taxon in each sample. In addition, a 300-count random subsample was taken using the `rarifyCounts()` function. In coarse mesh samples, a scan was used to identify large and rare taxa. No counts were recorded for these taxa, so they were used only for taxa richness metrics and only for the original counts (not 300-organism subsampling). Then the original and 300-organism fine and coarse mesh sample data were combined into a new sample type (ZONW). Distinctness of each taxon in each sample and sample type was determined using the `assignDistinct()` function. Non-native status was then determined for all taxa on a site-specific basis. 

  For the purposes of calculations, the metrics were broken down into several subsets, depending on which autecology traits were required to calculate them. There are also separate functions to calculate the regional multimetric indices (MMIs) (component and overall scores and condition classes) based on metric inputs. This vignette describes how to use the various functions along with the example datasets provided and the included taxa lists, to obtain the full set of metrics, a subset of metrics based on metric type, and the benthic multimetric index used in assigning biological condition.  

  The functions in this package allow for a great deal of flexibility in terms of input variable names and the taxalist used in calculating metrics. However, the actual values of various traits are assumed to match those used in NLA. This means that some metrics will not be calculated correctly if the taxalist used in the function does not use the same trait values. In addition, the functions assume that data have already been summarized to the level of sample and taxon, and a within-sample distinctness (0/1) already assigned to each taxon, so any steps required to get to this point must be done prior to use of the functions. All taxa in the count data must be included in the taxa list in order to be included in the calculations. Functions will look for specific types of inconsistencies, like missing variables, and either provide a warning or will fail to run, depending on the issue.    

  There are two functions that are used to process the initial raw data for coarse- and fine-mesh net samples. The first, `convertZoopCts_NLA()`, uses count data, tow volume, the concentrated sample volume, the volume counted, and the biomass factor. This function is used to calculate density and biomass for each taxon with the original and 300-organism count data. The second function, `prepZoopCombCts_NLA`, combines the data from the coarse- and fine-mesh net samples. Once these functions are run, the data are ready for metric calculations. All three sample types (ZOCN, ZOFN, ZONW) are necessary to calculate the full set of metrics. Metrics of all types are ultimately used in the regional MMIs, so it is easiest to calculate the full set to obtain those necessary for calculating MMIs.   


********
# 2 Prepare Data for Metric Calculation

In this example, the initial raw data for the two sample types (ZOCN and ZOFN) are already combined with the tow volume for each sample into a single dataset. From there, we need to combine the data to create the ZONW sample. For this, we start with the function `convertZoopCts_NLA()`, create the 300-organism subsamples using `rarifyCounts()`, and then combine the ZOFN and ZOCN samples using `prepZoopCombCts_NLA()`.

## 2.1 Calculate density and biomass (original counts)

First run for the original counts. Convert counts to biomass and densities. Then create combined sample (ZONW) for original samples. To this data, we need to add the IS_DISTINCT variable to identify distinct taxa for richness-related metrics. We do this using the `assignDistinct()` function. We need to first remove any taxa without merge the dataset with the taxalist in order to add taxonomic hierarchical information. This taxonomic hierarchy is used to determine whether each taxon is unique within a sample. We will leave the additional taxonomic information in the data frame for now. 

*Note:* For one UID, the ZOCN sample contained 0 organisms, denoted by the TAXA_ID 9999. Because the COUNT for this row is 0, we exclude it when we assign distinctness. 

```{r convertRaw-1, message=FALSE, warning=FALSE}
library(aquametBio)

head(zpRawEx)

zpCts.full <- convertZoopCts_NLA(zpRawEx, sampID='UID', sampType='SAMPLE_TYPE',
                            rawCt = 'ABUNDANCE_TOTAL', biofactor = 'BIOMASS_FACTOR',
                            tow_vol = 'TOW_VOLUME', vol_ctd = 'VOLUME_COUNTED',
                            conc_vol = 'CONCENTRATED_VOLUME', lr_taxon = 'LARGE_RARE_TAXA',
                            taxa_id = 'TAXA_ID', subsample=FALSE)

zpCts.full <- subset(zpCts.full, COUNT!=0|LARGE_RARE_TAXA=='Y') |>
  merge(zoopTaxa, by = 'TAXA_ID') |>
  assignDistinct(sampleID = c('UID', 'SAMPLE_TYPE'), 
                 taxlevels=c('PHYLUM','CLASS','SUBCLASS',
                             'ORDER','SUBORDER','FAMILY','GENUS',
                             'SPECIES','SUBSPECIES')) 

head(zpCts.full)
```

## 2.2 Create ZONW sample

Then create combined sample (ZONW) from the original coarse and fine net samples. We again need to add the IS_DISTINCT variable, using the same approach as above. 

```{r createZONW, message=FALSE, warning=FALSE}
zpCts.zonw <- prepZoopCombCts_NLA(zpCts.full, sampID = c('UID'),
                                  sampType = 'SAMPLE_TYPE', typeFine = 'ZOFN',
                                  typeCoarse = 'ZOCN', ct = 'COUNT',
                                  biomass = 'BIOMASS', density = 'DENSITY', 
                                  taxa_id = 'TAXA_ID', lr_taxon = 'LARGE_RARE_TAXA')

zpCts.zonw <- subset(zpCts.zonw, COUNT!=0|LARGE_RARE_TAXA=='Y') |>
  merge(zoopTaxa, by = 'TAXA_ID') |>
  assignDistinct(sampleID = c('UID', 'SAMPLE_TYPE'), 
                 taxlevels=c('PHYLUM','CLASS','SUBCLASS',
                             'ORDER','SUBORDER','FAMILY','GENUS',
                             'SPECIES','SUBSPECIES')) 


head(zpCts.zonw)
```

## 2.3 Create random subsamples and prepare data

Now we can create the random subsamples from the original count data using `rarifyCounts()`, and then repeat the process from above for the 300-organism subsamples. We append the COUNT and BIOMASS variables with the subsample size (300) to distinguish these values from the original counts since we will combine them all

```{r rarify, message=FALSE, warning = FALSE}
zpRawEx.300 <- rarifyCounts(zpRawEx, sampID = c('UID', 'SAMPLE_TYPE'),
                            abund = 'ABUNDANCE_TOTAL',
                            subsize = 300,
                            seed = 1234) |>
  subset(ABUNDANCE_TOTAL > 0, select = -LARGE_RARE_TAXA)

head(zpRawEx.300)
```

Now that we have the 300 organism subsamples for coarse- and fine-mesh net samples, we can convert counts to biomass. We cannot use large and rare taxa for the 300-organism subsample and we do not use densities for the subsetted data, so these variables are missing. We also need to calculate IS_DISTINCT at this point using the `assignDistinct()` function. We add the suffix _300 to the variable name to distinguish it from original count data. 

```{r subsamp-1, message=FALSE, warning = FALSE}
zpCts.300 <- convertZoopCts_NLA(zpRawEx.300, sampID='UID', sampType='SAMPLE_TYPE',
                            rawCt = 'ABUNDANCE_TOTAL', biofactor = 'BIOMASS_FACTOR',
                            tow_vol = 'TOW_VOLUME', vol_ctd = 'VOLUME_COUNTED',
                            conc_vol = 'CONCENTRATED_VOLUME', 
                            taxa_id = 'TAXA_ID', subsample=TRUE) 

zpCts.300 <- subset(zpCts.300, COUNT!=0) |>
  merge(zoopTaxa, by = 'TAXA_ID') |>
  assignDistinct(sampleID = c('UID', 'SAMPLE_TYPE'), 
                 taxlevels=c('PHYLUM','CLASS','SUBCLASS',
                             'ORDER','SUBORDER','FAMILY','GENUS',
                             'SPECIES','SUBSPECIES')) |>
  subset(select = c('UID', 'SAMPLE_TYPE', 'TAXA_ID', 'COUNT', 'BIOMASS', 'IS_DISTINCT'))

names(zpCts.300)[names(zpCts.300) %in% c('COUNT', 'BIOMASS', 'IS_DISTINCT')] <- paste(names(zpCts.300)[names(zpCts.300) %in% c('COUNT', 'BIOMASS', 'IS_DISTINCT')], "300", sep="_") 

head(zpCts.300)
```

Finally, we combine the coarse and fine net samples into the ZONW sample type for the 300-count data. After combining, we need to calculate IS_DISTINCT as above. Again, we add the suffix _300 to the name to distinguish it from original count data. Again, only keep necessary variables because we will be merging with the broader original count data, which contains non-native status information.

```{r subsamp-comb, message=FALSE, warning = FALSE}
zpCts.zonw.300 <- prepZoopCombCts_NLA(zpCts.300, sampID = c('UID'),
                                  sampType = 'SAMPLE_TYPE', typeFine = 'ZOFN',
                                  typeCoarse = 'ZOCN', ct = 'COUNT_300',
                                  biomass = 'BIOMASS_300', 
                                  taxa_id = 'TAXA_ID')

zpCts.zonw.300 <- subset(zpCts.zonw.300, COUNT_300!=0) |>
  merge(zoopTaxa, by = 'TAXA_ID') |>
  assignDistinct(sampleID = c('UID', 'SAMPLE_TYPE'), 
                 taxlevels=c('PHYLUM','CLASS','SUBCLASS',
                             'ORDER','SUBORDER','FAMILY','GENUS',
                             'SPECIES','SUBSPECIES')) |>
  subset(select = c('UID', 'SAMPLE_TYPE', 'TAXA_ID', 'COUNT_300', 'BIOMASS_300', 'IS_DISTINCT'))

names(zpCts.zonw.300)[names(zpCts.zonw.300) %in% c('IS_DISTINCT')] <- 'IS_DISTINCT_300'

head(zpCts.zonw.300)

```

## 2.4 Create inputs for metric calculation

We have converted counts to biomass and density for both the original and 300-organism subsamples, and combined the coarse (ZOCN) and fine (ZOFN) samples into a new sample type (ZONW). Now we need to combine all of the data into a single data frame for each sample type (ZOCN, ZOFN, and ZONW).

```{r combine, message=FALSE, warning = FALSE}
zpCts.zonw.all <- merge(zpCts.zonw, zpCts.zonw.300, 
                   by = c('UID', 'SAMPLE_TYPE', 'TAXA_ID'), all.x=TRUE)


head(zpCts.zonw.all)

zpCts.full.zocn <- subset(zpCts.full, SAMPLE_TYPE=='ZOCN')
zpCts.300.zocn <- subset(zpCts.300, SAMPLE_TYPE=='ZOCN')

zpCts.zocn.all <- merge(zpCts.full.zocn, zpCts.300.zocn, 
                  by = c('UID', 'SAMPLE_TYPE', 'TAXA_ID'), all.x=TRUE)

head(zpCts.zocn.all)

zpCts.full.zofn <- subset(zpCts.full, SAMPLE_TYPE=='ZOFN')
zpCts.300.zofn <- subset(zpCts.300, SAMPLE_TYPE=='ZOFN')

zpCts.zofn.all <- merge(zpCts.full.zofn, zpCts.300.zofn, 
                  by = c('UID', 'SAMPLE_TYPE', 'TAXA_ID'), all.x=TRUE)

head(zpCts.zofn.all)
```

## 2.5 Assign native status to taxa

We have the NON_NATIVE variable from the taxalist now included in the three input files for metric calculations. We will create a data frame that contains the state and the ecoregion of each UID. The state variable is needed to assign non-native status based on the way the NON_NATIVE variable exists in the taxalist (zoopTaxa). The ecoregion variable (ECO_BIO) variable will be used later to calculate the MMI. After we add NON_NATIVE_TAXON to each input dataset with values of 1 for non-native and 0 for native, we can select only the variables we need for metric calculation.

```{r non-native, message=FALSE, warning = FALSE}
states <- data.frame(UID = c(7100, 7122, 7477, 7548, 7748, 8395, 8499, 
                             2010287, 2010569, 2010833),
                     STATE = c('OR', 'PA', 'ID', 'IN', 'MN', 'UT', 'IL', 'MI', 
                               'SD', 'SC'),
                     ECO_BIO = c('WMTNS', 'EHIGH', 'WMTNS', 'EHIGH', 'UMW',
                                 'WMTNS', 'PLAINS', 'UMW', 'PLAINS', 'EHIGH'))

# Combined sample
zonwIn.nn <- merge(zpCts.zonw.all, states, by = 'UID') 

zonwIn.nn$NON_NATIVE_TAXON <- with(zonwIn.nn, 
                                          ifelse(is.na(NON_NATIVE), 0,
                                            ifelse(NON_NATIVE=='ALL', 1,
                                              ifelse(grep(STATE, NON_NATIVE), 1, 0)))) 

zonwIn.nn <- zonwIn.nn[, c('UID', 'SAMPLE_TYPE', 'TAXA_ID', 
                                           'COUNT', 'BIOMASS', 'DENSITY',
                                           'IS_DISTINCT', 'LARGE_RARE_TAXA',
                                           'COUNT_300', 'BIOMASS_300',
                                           'IS_DISTINCT_300',
                                           'ECO_BIO', 'NON_NATIVE_TAXON')]

head(zonwIn.nn)

# Coarse mesh net sample
zocnIn.nn <- merge(zpCts.zocn.all, states, by = 'UID') 

zocnIn.nn$NON_NATIVE_TAXON <- with(zocnIn.nn, 
                                          ifelse(is.na(NON_NATIVE), 0,
                                            ifelse(NON_NATIVE=='ALL', 1,
                                              ifelse(grep(STATE, NON_NATIVE), 1, 0)))) 

zocnIn.nn <- zocnIn.nn[, c('UID', 'SAMPLE_TYPE', 'TAXA_ID', 
                                           'COUNT', 'BIOMASS', 'DENSITY',
                                           'IS_DISTINCT', 'LARGE_RARE_TAXA',
                                           'COUNT_300', 'BIOMASS_300',
                                           'IS_DISTINCT_300',
                                           'ECO_BIO', 'NON_NATIVE_TAXON')]

head(zocnIn.nn) 

# Fine mesh net sample
zofnIn.nn <- merge(zpCts.zofn.all, states, by = 'UID') 

zofnIn.nn$NON_NATIVE_TAXON <- with(zofnIn.nn, 
                                          ifelse(is.na(NON_NATIVE), 0,
                                            ifelse(NON_NATIVE=='ALL', 1,
                                              ifelse(grep(STATE, NON_NATIVE), 1, 0)))) 

zofnIn.nn <- zofnIn.nn[, c('UID', 'SAMPLE_TYPE', 'TAXA_ID', 
                                           'COUNT', 'BIOMASS', 'DENSITY',
                                           'IS_DISTINCT', 'LARGE_RARE_TAXA',
                                           'COUNT_300', 'BIOMASS_300',
                                           'IS_DISTINCT_300',
                                           'ECO_BIO', 'NON_NATIVE_TAXON')]

head(zofnIn.nn)
```

# 3 Calculate Metrics

Because of the complexity of the full set of metrics, there are several functions that can be used to calculate subsets of metrics. To use the functions that run individual sets of metrics, we need to run for each set of data (original data and based on the 300-organism subsamples), and we need to account for native status. The process is somewhat cumbersome, so there is also a function that just calculates all of the metrics that were evaluated for NLA at once. We will first show a run of a few of the functions that calculate subsets of the data.

## 3.1  Calculate base metrics

Base metrics are a large set of metrics based on taxonomic, trait information or a combination of the two. This would need to be run for each combination desired: 1) original data, 2) original native only, 3) 300-organism subsample data, and 4) 300-organism subsample native only. In this example, we will only show the original native taxa only and the 300-organism subsample with all taxa. The output from the metric function is in the form of: sample ID variables (here UID), PARAMETER, and RESULT. Thus, to show the metrics calculated, we show the unique PARAMETER values in each data frame. 

For base metrics, only the combined data (coarse + fine) are used. Not that for subsample data, only counts and biomass are available for calculating metrics. 

### 3.1.1 Full counts with only native taxa

```{r calc-mets-1.1, message=FALSE, warning = FALSE}
# Full counts with native taxa only
zoopMets.base.nat <- calcZoopBaseMetrics(indata = zonwIn.nn, 
                                         sampID = c('UID', 'SAMPLE_TYPE'), 
                                         is_distinct = 'IS_DISTINCT',
                                         ct = 'COUNT', 
                                         biomass = 'BIOMASS', 
                                         density = 'DENSITY',
                                         inTaxa = zoopTaxa, 
                                         taxa_id='TAXA_ID',
                                         ffg = 'FFG', 
                                         clad_size = 'CLADOCERA_SIZE',
                                         net_size = 'NET_SIZECLS_NEW',
                                         nativeMetrics = TRUE, 
                                         nonnative = 'NON_NATIVE_TAXON')

noquote(unique(zoopMets.base.nat$PARAMETER))
```

### 3.1.2 Subsample counts with all taxa

```{r calc-mets-1.2, message=FALSE, warning = FALSE}
# 300-organism subsamples
zoopMets.300 <- calcZoopBaseMetrics(indata = zonwIn.nn, 
                                    sampID = c('UID', 'SAMPLE_TYPE'), 
                                    is_distinct = 'IS_DISTINCT_300',
                                    ct = 'COUNT_300', 
                                    biomass = 'BIOMASS_300',
                                    inTaxa = zoopTaxa, 
                                    taxa_id='TAXA_ID',
                                    ffg = 'FFG', 
                                    clad_size = 'CLADOCERA_SIZE',
                                    net_size = 'NET_SIZECLS_NEW',
                                    nativeMetrics = FALSE)

noquote(unique(zoopMets.base.nat$PARAMETER))
```
## 3.2 Calculate diversity metrics

Diversity metrics are again calculated for each subset of data desired, with the appropriate inputs. For these metrics, the combined ZONW sample is used. For the full data set, the process is straightforward. For subsampled data, the names of the output metrics must be modified manually. To run diversity metrics on specific taxonomic groups, the input data must already be subset, but suffix can be specified and will be added automatically to parameter names. 

### 3.2.1 Using full counts but native taxa only

```{r calc-mets-2.1, message=FALSE, warning = FALSE}
# Full counts with native taxa only
zoopMets.div <- calcZoopDivMetrics(zonwIn.nn, 
                                        sampID = 'UID', 
                                        is_distinct = 'IS_DISTINCT',
                                        ct = 'COUNT', 
                                        biomass = 'BIOMASS', 
                                        density = 'DENSITY')

noquote(unique(zoopMets.div$PARAMETER))

```

### 3.2.2 Using 300-organism subsamples, all taxa

```{r calc-mets-2.2, message=FALSE, warning = FALSE}
# 300-organism subsamples
zoopMets.div.300 <- calcZoopDivMetrics(zonwIn.nn, 
                                       sampID = 'UID', 
                                       is_distinct = 'IS_DISTINCT_300',
                                       ct = 'COUNT_300', 
                                       biomass = 'BIOMASS_300')

# Update parameter names to reflect subsample
zoopMets.div.300$PARAMETER <- gsub("\\_NIND", "300\\_NIND", zoopMets.div.300$PARAMETER)
zoopMets.div.300$PARAMETER <- gsub("\\_BIO", "300\\_BIO", zoopMets.div.300$PARAMETER)

noquote(unique(zoopMets.div.300$PARAMETER))
```

### 3.2.3 Using full counts but only rotifers

```{r calc-mets-2.3, message=FALSE, warning = FALSE}
# Rotifer diversity
# First need to merge taxa list with data to be able to select rotifers only
zonwIn.taxa <- merge(zonwIn.nn, zoopTaxa, by = 'TAXA_ID') |>
    subset(select = c('UID', 'TAXA_ID', 'COUNT', 'COUNT_300', 'BIOMASS', 
                      'BIOMASS_300', 'DENSITY', 'IS_DISTINCT', 
                      'IS_DISTINCT_300', 'SUBORDER', 'SUBCLASS', 'PHYLUM'))

# Calculate rotifer diversity for full counts
zoopMets.div.rot <- calcZoopDivMetrics(subset(zonwIn.taxa, PHYLUM=='ROTIFERA'),
                                       sampID = 'UID', 
                                       is_distinct = 'IS_DISTINCT',
                                       ct = 'COUNT', 
                                       suffix = 'ROT')

noquote(unique(zoopMets.div.rot$PARAMETER))
```

### 3.2.4 Using subsample data but only rotifers

```{r calc-mets-2.4, message=FALSE, warning = FALSE}
# Calculate rotifer diversity for subsample data
zoopMets.div.rot.300 <- calcZoopDivMetrics(subset(zonwIn.taxa, PHYLUM=='ROTIFERA'),
                                       sampID = 'UID', 
                                       is_distinct = 'IS_DISTINCT_300',
                                       ct = 'COUNT_300', 
                                       suffix = 'ROT')
# Update parameter names to reflect subsample
zoopMets.div.rot.300$PARAMETER <- paste0(zoopMets.div.rot.300$PARAMETER, '300')

noquote(unique(zoopMets.div.rot.300$PARAMETER))
```

## 3.3 Calculate richness metrics

For richness metrics, we use all three datasets created based on sample type (coarse-mesh, fine-mesh, and combined). The main metrics come from the combined dataset (ZONW), and others based on data from a single mesh size (ZOCN and ZOFN) have modified names. In all of these examples, a version of TOTL_NTAX is included in the output dataset. Normally, if we were combining these results with those from other datasets, these would be removed from the output because they are calculated elsewhere. They have been left in the output here for demonstration purposes. 

### 3.3.1 Full and subsample counts - combined sample

```{r calc-mets-3.1, message = FALSE, warning = FALSE}
zoopMets.zonw <- calcZoopRichnessMetrics(zonwIn.nn, 
                                         sampID = 'UID',
                                         distVars=c('IS_DISTINCT', 'IS_DISTINCT_300'),
                                         nonnative = 'NON_NATIVE_TAXON', 
                                         inTaxa = zoopTaxa,
                                         taxa_id = 'TAXA_ID', 
                                         genus = 'GENUS',
                                         family = 'FAMILY', 
                                         prefix = c('', '300')) 

noquote(unique(zoopMets.zonw$PARAMETER))
    
```

### 3.3.2 Full and subsample counts - coarse-mesh sample data

```{r calc-mets-3.2, message = FALSE, warning = FALSE}
  zoopMets.zocn <- calcZoopRichnessMetrics(zocnIn.nn, 
                                             sampID = 'UID',
                                             distVars=c('IS_DISTINCT', 'IS_DISTINCT_300'),
                                             nonnative = 'NON_NATIVE_TAXON', 
                                             inTaxa = zoopTaxa,
                                             taxa_id = 'TAXA_ID', 
                                             genus = 'GENUS',
                                             family = 'FAMILY', 
                                             prefix = c('', '300'))

  zoopMets.zocn$PARAMETER <- gsub('FAM\\_', 'ZOCN\\_FAM\\_', zoopMets.zocn$PARAMETER)
  zoopMets.zocn$PARAMETER <- gsub('GEN\\_', 'ZOCN\\_GEN\\_', zoopMets.zocn$PARAMETER)
  zoopMets.zocn$PARAMETER <- gsub('FAM300', 'ZOCN300\\_FAM',
                                    zoopMets.zocn$PARAMETER)
  zoopMets.zocn$PARAMETER <- gsub('GEN300', 'ZOCN300\\_GEN',
                                    zoopMets.zocn$PARAMETER)
  
  noquote(unique(zoopMets.zocn$PARAMETER))
```


## 3.4 Calculate all metrics

Above were just two examples of using the functions to calculate subsets of the zooplankton metrics. The easiest approach to calculating all of the metrics is to use the function `calcZoopAllMets()`. It requires both original and subsamples counts, as well as a native status designation for each taxon in the count data, with values of 0 = native and 1 = non-native. 

```{r calc-mets-all, message=FALSE, warning=FALSE}
zoopMets <- calcZoopAllMets(indata = zonwIn.nn,
                               inCoarse = zocnIn.nn,
                               inFine = zofnIn.nn,
                               inTaxa = zoopTaxa,
                               sampID = c('UID'),
                               is_distinct = 'IS_DISTINCT',
                               ct = 'COUNT', biomass = 'BIOMASS',
                               density = 'DENSITY',
                               is_distinct_sub = 'IS_DISTINCT_300',
                               ct_sub = 'COUNT_300',
                               biomass_sub = 'BIOMASS_300',
                               sub_mod = '300', taxa_id = 'TAXA_ID',
                               nonnative = 'NON_NATIVE_TAXON', genus = 'GENUS',
                               family = 'FAMILY', ffg = 'FFG',
                               clad_size = 'CLADOCERA_SIZE',
                               net_size = 'NET_SIZECLS_NEW')

head(zoopMets)

noquote(unique(zoopMets$PARAMETER))

```

# 4 Calculate MMI and assign condition

Now that we have the full set of metrics, we can calculate the MMI and assign a condition class to each sample.  To determine which MMI to apply to each site, we need the aggregated ecoregion (ECO_BIO) as used in the National Lakes Assessment (NLA). We created a data frame with this information in section 2.5 where native status was assigned. Thus, we will start by merging the `states` data frame with the output from section 3.4, then reshaping the data frame wide because that is what is expected in the `calcNLA_ZoopMMI` function. 

## 4.1 Calculate MMI

```{r calc-MMI, message = FALSE, warning = FALSE}
mmiIn <- merge(states, zoopMets, by = 'UID') |>
  reshape(idvar = c('UID', 'ECO_BIO'), direction = 'wide',
                             timevar = 'PARAMETER', v.names = 'RESULT')

names(mmiIn) <- gsub('RESULT\\.', '', names(mmiIn))

mmiOut <- calcNLA_ZoopMMI(inMets = mmiIn, 
                          sampID = 'UID',
                          ecoreg = 'ECO_BIO',
                          totlnind = 'TOTL_NIND')

head(mmiOut)

```
